Staged Abstract Interpreters

### 2019-04-05 paper/10_intro.tex
paper

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and emprical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    \note{garbage collection}, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive, \note{context-sensitive, and store-widening} analysis.
\end{itemize}

>>>>

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while untouch 
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We discuss the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    \note{garbage collection}, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive, \note{context-sensitive, and store-widening} analysis.
\end{itemize}

### 2019-04-05 paper/70_eval.tex
paper

To evaluate the performance can be improved by our staging abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analyzing time against with the unstaged versions. The result
shows an average \todo{X}\% running time improvement, as a strong evidence of
our \textit{abstraction without regret} approach to construct abstract
interpreters.

We implement the abstract interpreters for control-flow analysis, i.e., a
value-flow analysis focus on lambda terms in functional languages. A frontend
desugars Scheme to a small core language that make the analyzer easy to
implement. Macros and @call/cc@ is not handled. As described in Section
\ref{cfa}, we implement a monovariant 0-CFA-like analysis, a context-sensitive
$1$-CFA-like analysis, both their versions with and without store-widening.

\paragraph{Setting}
We use Scala 2.12.7 and Oracle JVM 1.8.0-191, running on an Ubuntu 16.04 LTS
(kernel 4.4.0) machine. All of our evaluation were performed on a machine with 4 Intel
Xeon Platinum 8168 CPU at 2.7GHz and 3 TiB of RAM. Although the machine has 96
cores (192 threads) in total, the abstract interpreters are single-threaded.
To minimize the effect of warming-up from the JVM JIT compiler, all the
experiments are executed for 20 times and we report the statistical median value
of the running times (in seconds).

\paragraph{Benchmarks}
The benchmark programs we used in the experiment are collected from several previous papers
\cite{Johnson:2013:OAA:2500365.2500604, ashley:practical,
DBLP:journals/corr/abs-1102-3676} and exisiting artifacts
\footnote{https://github.com/ilyasergey/reachability} for control-flow analysis.
Some of the benchmarks are small programs designed for experiments, such as the @kcfa-worst-@$n$ series, 
which are intended to challenge $k$-CFA; while some other benchmarks are from real-world applications, 
for examples, the RSA public key encryption algorithm @rsa@ and matrix computation @matrix@.
In the Table \ref{?}, we report the number of AST node after parsing and desugaring 
(excluding comments) as a proper measurement of program size.

@@

Figure~\ref{evaluation_result} shows the evaluation results. The
\textit{unstaged} and \textit{staged} columns show the median time to finish the
analysis (in seconds). The "$\Delta$" column shows the median runtime change
from the unstaged to the staged version. \todo{0cfa, 1cfa, store-widen}

>>>>

To evaluate the performance can be improved by our staging abstract interpreters
approach, we scale our toy interpreter to a large subset of the Scheme language
and evaluate the analyzing time against with the unstaged versions. The result
shows that the staged versions are averagely 6 times as fast of the unstaged
version, as a strong evidence of our \textit{abstraction without regret}
approach.

\paragraph{Implementation and Evaluation Environment}
We implement the abstract interpreters for control-flow analysis, i.e., a
value-flow analysis focus on lambda terms in functional languages. A frontend
desugars Scheme to a small core language that make the analyzer easy to
implement. As described in Section \ref{cfa}, we implement a monovariant
0-CFA-like analysis, \todo{a context-sensitive $1$-CFA-like analysis, both their
  versions with and without store-widening}. We also implement several
optimizations mentioned in Section \ref{staged_ds} (less than ~30 LOC in total).

We use Scala 2.12.7 and Oracle JVM 1.8.0-191, running on an Ubuntu 16.04 LTS
(kernel 4.4.0) machine. All of our evaluation were performed on a machine with 4 Intel
Xeon Platinum 8168 CPU at 2.7GHz and 3 TiB of RAM. Although the machine has 96
cores (192 threads) in total, the abstract interpreters are single-threaded.
To minimize the effect of warming-up from the JVM JIT compiler, all the
experiments are executed for 20 times and we report the statistical median value
of the running times (in seconds).

\paragraph{Benchmarks}
The benchmark programs we used in the experiment are collected from several
previous papers \cite{Johnson:2013:OAA:2500365.2500604, ashley:practical,
DBLP:journals/corr/abs-1102-3676} and exisiting artifacts
\footnote{https://github.com/ilyasergey/reachability} for control-flow analysis.
Some of the benchmarks are small programs designed for experiments, such as the
@kcfa-worst-@$n$ series, which are intended to challenge $k$-CFA; while some
other benchmarks are from real-world applications, for examples, the RSA public
key encryption algorithm @rsa@. In the Figure \ref{evaluation_result}, we report
the number of AST node after parsing and desugaring (excluding comments) as a
proper measurement of program size.

@@

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on 0CFA without store-widening. The
\textit{unstaged} and \textit{staged} columns show the median time to finish the
analysis (in seconds). The last column $\frac{\text{unstaged}}{\text{staged}}$
shows the improvement from staging. \todo{0cfa, 1cfa, store-widen}
As we can see from the table, the staged version significantly outperforms the
unstaged version on all benchmarks, especially for the @kcfa-worst-@$n$ series.
We observe the most improvement of 14x times faster than the unstaged version on
@kcfa-worst-64@. On average, the staging produces code that run 6x times faster
than the unstaged analyzer.

We also observe that much overhead of unstaged analyzers are from the monadic
layers. Thus our evaluation result provides an abstraction without regret
approach to construct optimizing abstract interpreters -- user may write an
high-level modular and composable abstract interpreter, and then derive a staged
version in a principled way that runs significantly faster.

### 2019-04-05 paper/80_related.tex
paper

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to construct sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17}.
Similar to monadic abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present an arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\cite{DBLP:conf/cc/CousotC02} proposed an
abstract interpretation framework for modular analysis.
\citet{DBLP:conf/popl/CalcagnoDOY09} developed compositional shape analysis by
using bi-abduction.

\paragraph{Meta-Programming} The Lightweight Modular Staging framework relies on
types as stage annotations, and the staging procedure is modular. Other notable
implementations of MSP exist in ML family, for examples, MetaML
\cite{DBLP:conf/pepm/TahaS97} and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03,
DBLP:conf/flops/Kiselyov14}. Compared with the LMS approach in Scala,
MetaML/MetaOCaml use term-level annotations such as brackets, escape, and run.
Notwithstanding, we use LMS in this paper, the idea of staging an abstract
interpreter still applies with other MSP implementations. Partial evaluation as
an automatic technique is studied comprehensively
\cite{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}. Futamura projections
reveals close relations between interpreters and compilers \cite{Futamura1999,
futamura1971partial}. \citet{DBLP:conf/gpce/Asai14} studies the compilation of
reflective language using MSP and MetaOCaml.
\citeauthor{Amin:2017:CTI:3177123.3158140} shows how to collapse a tower of
concrete interpreters by MSP and LMS framework.

@@

MSP with functors and monads \cite{10.1007/11561347_18}

staged monadic interpreters \cite{DBLP:conf/dsl/SheardBP99}

compiling monads \cite{danvy1991compiling}

>>>>

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to construct sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
Similar to monadic abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis.

\paragraph{Meta-Programming} In this paper, we use the Lightweight Modular
Staging framework \cite{DBLP:conf/gpce/RompfO10}, which relies on types as stage
annotations. Other notable implementations of MSP exist in ML family, e.g.,
MetaML \cite{DBLP:conf/pepm/TahaS97} and MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
The idea of staging an abstract interpreter we presented in this paper is also
applicable to other MSP system. Partial evaluation as an automatic technique is
studied comprehensively \cite{10.1007/3-540-61580-6_11,
DBLP:books/daglib/0072559}. Futamura projections reveals close relations between
interpreters and compilers, which was originally proposed by
\citeauthor{futamura1971partial} in 1970s\cite{futamura1971partial}, and later
reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} shows how to collapse a tower of
concrete interpreters by MSP -- it would be also interesting to explore multiple
layers of abstract interpreters.
\citet{10.1007/11561347_18} discussed combining
multi-stage programming with functors and monads.
Similar to the idea in this paper, specializing monadic interpreters was
explored by \citet{DBLP:conf/dsl/SheardBP99, danvy1991compiling}.

@@
